Estimating length of SARS-CoV-2 from doubly interval censored data with missed infections

### 2024-03-20 cis-imperfect-testing.tex
Fix refs

Several mechanisms for episodes being undetected were considered when deriving $p_{iu}$ in \cref{E-perf-test:eq:piu}.
I now consider the additional mechanisms arising due to false negatives.
Specifically, episode $i$ could be undetected if the first test after the $b_{j(i)}$, episode $i$'s start day, is a false negative and then there are no subsequent positive tests.

@@

Therefore, this mechanism causes episode $i$ to be undetected if all the following conditions hold.
\begin{enumerate}
    \item The episode would have been detected considering only the mechanisms in \cref{E-perf-test:eq:piu}. That is $b_{j(i)} > \min(\sched_i)$ and $\tau_{\sched_i}(b_{j(i)}) + b_{j(i)} \leq e_{j(i)}$.
    \item The episode ends in the interval $[\tau_{\sched_i}(b_{j(i)}) + b_{j(i)}, \tau^2_{\sched_i}(b_{j(i)}) + b_{j(i)}]$.
      Note that the lower bound here is exactly the bound on $e_{j(i)}$ in the previous condition.
      Equivalently, $\tau_{\sched_i}(b_{j(i)}) + 1 \leq d_i \leq \tau^2_{\sched_i}(b_{j(i)}) + 1$.
    \item A false negative occurs on day $\tau_{\sched_i}(b_{j(i)}) + b_{j(i)}$. Conditional on the previous condition, this occurs with probability $1 - \psens$.
\end{enumerate}

The probability of this occurring, conditional on $B_{j(i)} = b_{j(i)} > \min(\sched_i)$ is:
\begin{align}
&\prob \left(
    \tau_{\sched_i}(b_{j(i)}) + 1 \leq D_i \leq \tau^2_{\sched_i}(b_{j(i)}) + 1
    \mid b_{j(i)},
\theta \right) (1 - \psens) \\
&= \left( S_\theta(\tau_{\sched_i}(b_{j(i)}) + 1) - S_\theta(\tau^2_{\sched_i}(b_{j(i)}) + 1) \right) (1 - \psens).
\end{align}
Integrating over $b_{j(i)}$ for $\min \sched_i < b_{j(i)} \leq T$ ($T$ being the last day of the period considered), in the same way as \cref{E-perf-test:eq:piu}, gives:
\begin{align}
\eta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right).
\end{align}

>>>>

Several mechanisms for episodes being undetected were considered when deriving $p_{iu}$ in \cref{perf-test:eq:piu}.
I now consider the additional mechanisms arising due to false negatives.
Specifically, episode $i$ could be undetected if the first test after the $b_{j(i)}$, episode $i$'s start day, is a false negative and then there are no subsequent positive tests.

@@

Therefore, this mechanism causes episode $i$ to be undetected if all the following conditions hold.
\begin{enumerate}
    \item The episode would have been detected considering only the mechanisms in \cref{perf-test:eq:piu}. That is $b_{j(i)} > \min(\sched_i)$ and $\tau_{\sched_i}(b_{j(i)}) + b_{j(i)} \leq e_{j(i)}$.
    \item The episode ends in the interval $[\tau_{\sched_i}(b_{j(i)}) + b_{j(i)}, \tau^2_{\sched_i}(b_{j(i)}) + b_{j(i)}]$.
      Note that the lower bound here is exactly the bound on $e_{j(i)}$ in the previous condition.
      Equivalently, $\tau_{\sched_i}(b_{j(i)}) + 1 \leq d_i \leq \tau^2_{\sched_i}(b_{j(i)}) + 1$.
    \item A false negative occurs on day $\tau_{\sched_i}(b_{j(i)}) + b_{j(i)}$. Conditional on the previous condition, this occurs with probability $1 - \psens$.
\end{enumerate}

The probability of this occurring, conditional on $B_{j(i)} = b_{j(i)} > \min(\sched_i)$ is:
\begin{align}
&\prob \left(
    \tau_{\sched_i}(b_{j(i)}) + 1 \leq D_i \leq \tau^2_{\sched_i}(b_{j(i)}) + 1
    \mid b_{j(i)},
\theta \right) (1 - \psens) \\
&= \left( S_\theta(\tau_{\sched_i}(b_{j(i)}) + 1) - S_\theta(\tau^2_{\sched_i}(b_{j(i)}) + 1) \right) (1 - \psens).
\end{align}
Integrating over $b_{j(i)}$ for $\min \sched_i < b_{j(i)} \leq T$ ($T$ being the last day of the period considered), in the same way as \cref{perf-test:eq:piu}, gives:
\begin{align}
\eta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right).
\end{align}

### 2024-03-20 transmission.tex
Tweak final plots / captions
Fix refs

The second approach, which I call the \emph{prevalence-inferred} approach takes a two-step approach.
First, it uses the statistical model from \cref{E-biology-data:sec:MRP} to estimate prevalence.
Then, it transforms the posterior distribution of prevalence to a posterior distribution of incidence using the forward substitution as in \cref{E-inc-prev:eq:forward-substitute}.
At a high-level, the statistical model is:
\begin{align}
    \vec{P}  &\dist F_P \\
    \rho &\dist F_\rho \\
    Y_{i,t} \mid \vec{P}, \rho &\dist \BB(n_{i,t}, P_{i,t}, \rho)
\end{align}
where $F_P$ is analogous to $F_Z$ in providing a joint prior for the prevalence in each stratum and day combination, and all other notation is as in the incidence-inferred approach.
The first step is to fit this model, giving a posterior distribution for $\vec{P}$.

The second step is to transform the posterior distribution for $\vec{P}$ to a posterior distribution for $\vec{Z}$.
A posterior sample for $\vec{P}$ can be transformed to a posterior sample for $\vec{Z}$ based on \cref{E-inc-prev:eq:forward-substitute}, see \cref{backcalc:sec:methods} for details.

@@

Then, I transform this posterior sample using \cref{transmission:alg:calculate-Z}.
The algorithm is based on \cref{E-inc-prev:eq:forward-substitute}, which is repeated here for convenience:
\begin{align}
Z_t
% &= \frac{P_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
&= P_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
\label{transmission:eq:forward-substitute}
\end{align}
Dividing each side of \cref{transmission:eq:forward-substitute} by the total number of individuals in each stratum transforms the number of prevalent individuals to prevalence and incidence to incidence proportion.
The algorithm uses a deterministic relationship between incidence and prevalence within a stratum, as \cref{E-inc-prev:sec:observation-process} justifies; \ie, the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.

@@

\Cref{transmission:eq:forward-substitute} is derived from \cref{E-inc-prev:eq:EPt-matrix}: $\vec{P} = \matr{S} \vec{I}$.
For this to have a unique solution, $\vec{I}$ and $\vec{P}$ need the same dimension.
However, I am using a period of prevalence which starts after the first infections.
These prior infections will still contribute to the prevalence.
I assume that the incidence for $t \leq 1$ is constant.
Therefore, $P_1 = I_0 \sum_{i=1}^{\dmax} S(i)$ and $I_t = P_1 / \sum_{i=1}^{\dmax} S(i)$ for $t \leq 1$.

% First, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
% For each sample and stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
% I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.
% I solve this system using forward substitution, calculating the value of $Z_t$ in \cref{E-inc-prev:eq:forward-substitute} sequentially for $t = 1, \dots, T$.
% $t = 1$ is defined as the first day where prevalence data is used.
% \Cref{E-inc-prev:eq:forward-substitute} is repeated here for convenience:
% \begin{align}
% Z_t
% &= \frac{P_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
% &= P_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
% \label{transmission:eq:forward-substitute}
% \end{align}
% % \Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
% Dividing each side by the total number of individuals in each stratum transforms the number of prevalent individuals to prevalence and incidence to incidence proportion.

@@

The \emph{attack rate} is the proportion of the population that have been infected at least once.
In this model it is $1 - \vec{s}$, the proportion susceptible at the time the study ends.
The attack rate is highest in the $[11, 16)$ age group in the North West at 68\% (95\% CrI: 66--70\%) (see \cref{SEIR:fig:attack-rates}).
Some of these infections occurred prior to the period covered by this study; during the period of this study 23\% (95\% CrI: 21--24\%) of the population in the North West aged 11--16 was infected.
\begin{figure}
    \includegraphics{SEIR/CIS/attack_rates}
    \caption{Posterior median and 95\% CrI of the attack rate in each region and age group.}
    \todo[inline]{Maybe compare this to blood donors, as used in the prior.}
    \label{SEIR:fig:attack-rates}
\end{figure}

Overall, the model fits the data well (see \cref{SEIR:fig:prev,SEIR:fig:prev-appendix}).
There are a few features of the data which are not captured by the model and the posterior estimate appears to be a compromise with the prevalence too low in one age group and too high in another.
For example, in the North East region, the 11--16 age group has its prevalence overestimated in October but the 16--25 age group has its prevalence underestimated.
\begin{figure}
    \thisfloatpagestyle{empty}
    \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_main}}
    \captionsetup{width=0.8\paperwidth}
    \captionof{figure}[SEIR prevalence goodness-of-fit]{%
        Goodness-of-fit to prevalence data by region and age for selected regions (see \cref{SEIR:fig:prev-appendix} for others).
        The central ribbon and line show the posterior median and 95\% CrI estimate for the proportion of the relevant stratum RT-PCR positive on each day.
        The lighter outer ribbon is the 95\% posterior predictive interval for the proportion of test results that are positive each week.
        The dots show the observed prevalence in CIS, aggregated by week.
        For a well-calibrated model, 95\% of the dots would be within the outer ribbon.
    }
    \label{SEIR:fig:prev}
\end{figure}

>>>>

The second approach, which I call the \emph{prevalence-inferred} approach takes a two-step approach.
First, it uses the statistical model from \cref{E-biology-data:sec:MRP} to estimate prevalence.
Then, it transforms the posterior distribution of prevalence to a posterior distribution of incidence using the forward substitution as in \cref{inc-prev:eq:forward-substitute}.
At a high-level, the statistical model is:
\begin{align}
    \vec{P}  &\dist F_P \\
    \rho &\dist F_\rho \\
    Y_{i,t} \mid \vec{P}, \rho &\dist \BB(n_{i,t}, P_{i,t}, \rho)
\end{align}
where $F_P$ is analogous to $F_Z$ in providing a joint prior for the prevalence in each stratum and day combination, and all other notation is as in the incidence-inferred approach.
The first step is to fit this model, giving a posterior distribution for $\vec{P}$.

The second step is to transform the posterior distribution for $\vec{P}$ to a posterior distribution for $\vec{Z}$.
A posterior sample for $\vec{P}$ can be transformed to a posterior sample for $\vec{Z}$ based on \cref{inc-prev:eq:forward-substitute}, see \cref{backcalc:sec:methods} for details.

@@

Then, I transform this posterior sample using \cref{transmission:alg:calculate-Z}.
The algorithm is based on \cref{inc-prev:eq:forward-substitute}, which is repeated here for convenience:
\begin{align}
Z_t
% &= \frac{P_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
&= P_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
\label{transmission:eq:forward-substitute}
\end{align}
Dividing each side of \cref{transmission:eq:forward-substitute} by the total number of individuals in each stratum transforms the number of prevalent individuals to prevalence and incidence to incidence proportion.
The algorithm uses a deterministic relationship between incidence and prevalence within a stratum, as \cref{E-inc-prev:sec:observation-process} justifies; \ie, the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.

@@

\Cref{transmission:eq:forward-substitute} is derived from \cref{inc-prev:eq:EPt-matrix}: $\vec{P} = \matr{S} \vec{I}$.
For this to have a unique solution, $\vec{I}$ and $\vec{P}$ need the same dimension.
However, I am using a period of prevalence which starts after the first infections.
These prior infections will still contribute to the prevalence.
I assume that the incidence for $t \leq 1$ is constant.
Therefore, $P_1 = I_0 \sum_{i=1}^{\dmax} S(i)$ and $I_t = P_1 / \sum_{i=1}^{\dmax} S(i)$ for $t \leq 1$.

% First, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
% For each sample and stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
% I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.
% I solve this system using forward substitution, calculating the value of $Z_t$ in \cref{inc-prev:eq:forward-substitute} sequentially for $t = 1, \dots, T$.
% $t = 1$ is defined as the first day where prevalence data is used.
% \Cref{inc-prev:eq:forward-substitute} is repeated here for convenience:
% \begin{align}
% Z_t
% &= \frac{P_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
% &= P_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
% \label{transmission:eq:forward-substitute}
% \end{align}
% % \Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
% Dividing each side by the total number of individuals in each stratum transforms the number of prevalent individuals to prevalence and incidence to incidence proportion.

@@

The \emph{attack rate} is the proportion of the population that have been infected at least once.
In this model it is $1 - \vec{s}$, the proportion susceptible at the time the study ends.
The attack rate is highest in the $[11, 16)$ age group in the North West at 68\% (95\% CrI: 66--70\%) (see \cref{SEIR:fig:attack-rates}).
Some of these infections occurred prior to the period covered by this study; during the period of this study 23\% (95\% CrI: 21--24\%) of the population in the North West aged 11--16 was infected.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics{SEIR/CIS/attack_rates}}
    \caption{Posterior median and 95\% CrI of the attack rate in each region and age group.}
    \todo[inline]{Maybe compare this to blood donors, as used in the prior.}
    \label{SEIR:fig:attack-rates}
\end{figure}

Overall, the model fits the data well (see \cref{SEIR:fig:prev,SEIR:fig:prev-appendix}).
There are a few features of the data which are not captured by the model and the posterior estimate appears to be a compromise with the prevalence too low in one age group and too high in another.
For example, in the North East region, the 11--16 age group has its prevalence overestimated in October but the 16--25 age group has its prevalence underestimated.
\begin{figure}
    % \thisfloatpagestyle{empty}
    % \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_main}}
    \captionsetup{width=0.8\paperwidth}
    \captionof{figure}[SEIR prevalence goodness-of-fit]{%
        Goodness-of-fit to prevalence data by region and age for selected regions (see \cref{SEIR:fig:prev-appendix} for others).
        The central ribbon and line show the posterior median and 95\% CrI estimate for the proportion of the relevant stratum RT-PCR positive on each day.
        The lighter outer ribbon is the 95\% posterior predictive interval for the proportion of test results that are positive each week.
        The dots show the observed prevalence in CIS, aggregated by week.
        For a well-calibrated model, 95\% of the dots would be within the outer ribbon.
    }
    \label{SEIR:fig:prev}
\end{figure}
